\jaxcosmo: End-to-End Differentiable and GPU Accelerated Cosmology Library

### 2022-05-30 paper.tex
CCIN2P3 acknowlegments
Tentative to write a short version of the Modeling & implementation section. The guideline is not to duplicate CCL article but to make clear what is in  the present version, what is not there that someone can think about (eg. neutrino contrib)...

\section{Modeling and implementation}
\label{sec-jaxcosmo-design}
In this section, we describe the cosmological modeling provided by jax-cosmo, and its implementation in JAX.

@@

\subsection{Background cosmology}
% Usual stuff

\subsection{Growth of perturbations}
% The interesting point to mention here is the differentiable growth ODE

\subsection{Matter power spectrum}
% Here we describe the EH power spectrum, mention possible extensions with emulators

\subsection{Tracers}
% Here we describe the kernels, like in CCL

\subsection{Angular power-spectra}
% Limber integration, this can be an opportunity to talk about differentiation through integrals as well.

%
\subsection{Validation against the Core Cosmology Library (CCL)}
To illustrate the different features available with the present version of the library, we have written a  notebook to make  comparison with the well established  Core Cosmology Library \citep{2019ApJS..242....2C}. As an example, the Figure \ref{fig:halofit_comparison} shows an example of the non linear matter power spectrum computation using the \texttt{halofit} fitting function \citep{2012ApJ...761..152T, 2003MNRAS.341.1311S}. 
%
\begin{figure}[ht]
    \centering
    \includegraphics[width=\columnwidth]{figures/halofit_pk.png}
    \caption{Comparison of matter power spectrum between CCL and jax-cosmo. \textbf{Example of validation plot} \nblink{CCL_comparison}}
    \label{fig:halofit_comparison}
\end{figure}

@@

%
%
\subsection{DES-Y1 Example}
%

@@

Then, using the \numpyro\ PPL, we have set up a \textit{forward model} using the prior description  Table \ref{tab-DESY1} as for instance the $\Omega_c$ parameter
\begin{minted}[fontsize=\footnotesize]{python}
Omega_c = numpyro.sample('Omega_c', Uniform(0.1, 0.9))
\end{minted}
Then, we generate mock angular power spectra $C_\ell$ from the auto \& cross correlations of the Number Counts and Weak Lensing probes computed with the \texttt{gaussian\_cl\_covariance\_and\_mean} function in the \texttt{jc.angular\_cl} module. One reads
\begin{minted}[fontsize=\footnotesize]{python}
cl, C = gaussian_cl_covariance_and_mean(cosmo, ell, probes, 
                               f_sky=0.25, sparse=True)
\end{minted} 
with \texttt{cosmo} an instance of the \texttt{jc.Cosmology} setting the cosmological parameters generated with the priors, and \texttt{ell} a series of 50 angular modes. The generation of the mock data yields from the possibility to execute a conditioned model with a fiducial cosmology, leaving \numpyro\ dealing with the random number generation:
\begin{minted}[fontsize=\footnotesize]{python}
fiducial_model = condition(model,
    {'Omega_c':0.2545, 'sigma8':0.801, 
    'h':0.682, 'Omega_b':0.0485, 'w0':-1.,'n_s':0.971,
     'A':0.5,'eta':0.,
     'm1':0.0,'m2':0.0,'m3':0.0,'m4':0.0,
     'dz1':0.0,'dz2':0.0,'dz3':0.0,'dz4':0.0,
     'b1':1.2,'b2':1.4,'b3':1.6,'b4':1.8,'b5':2.0
      })

>>>>

\section{Modeling and implementation}
\label{sec-jaxcosmo-design}
In this section, we describe the cosmological modeling provided by \jaxcosmo, and its implementation in JAX. The general development follows the Core Cosmology Library structure \citep{2019ApJS..242....2C} although it does not include all the options as the present \jaxcosmo\ version is more to be viewed as a proof of concept.

The cosmological parameters are organised in a JAX \textit{pyTree} container\footnote{\url{https://jax.readthedocs.io/en/latest/jax-101/05.1-pytrees.html}} which allows one to use \textit{autodiff} and \texttt{vmap} transparently. The present choice of parameters are: $\Omega_{cdm}$ the cold dark matter density fraction, $\Omega_b$ the Byronic matter density fraction, $h$ the Hubble constant divided by $100$~km/s/Mpc, $n_s$ the primordial scalar perturbation spectral index, $\sigma_8$ the variance of matter density perturbations at an $8$~Mpc/h scale, $\Omega_k$ the curvature density fraction, $w_0$ the first order term of dark energy equation and $w_a$ the second order term of dark energy equation of state.

The traditional computation of the different cosmological background follows \cite{2005A&A...443..819P}  with namely the Friedmann equation
\begin{equation}
    E^2(a) = \frac{H^2(a)}{H^2_0} = \Omega_m a^{-3} + \Omega_k a^{-2} + \Omega_{de} e^{f(a)}
\end{equation}
with $a=1/(1+z)$ the scale factor related to the redshift $z$, $H(a)=\dot{a}/a$ the Hubble parameter with $H_0$ its present day value, $\Omega_m= \Omega_{cdm}+\Omega_b$, $\Omega_{de}=1-\Omega_k-\Omega_m$, and 
\begin{equation}
    f(a) = -3 (1 + w_0 + w_a) \ln(a) + 3 w_a (a - 1)
\end{equation}
Noticeably, the relativist contributions of massless neutrinos and photon radiations, as well as the massive neutrino contribution are neglected.

Currently, \jaxcosmo\  implements the \cite{Eisenstein_1998} transfer function and non-linear features through the \textit{halofit} model by \cite{2012ApJ...761..152T} or \cite{2003MNRAS.341.1311S} (w/o the neutrino contribution introduced by \cite{10.1111/j.1365-2966.2011.20222.x}). It is worth to mention that to perform \textit{autodiff} we have implemented a series of tools as differentiable ODE (Romberg algorithm) as well as integration quadrature (eg. Simpson).

Future version of the library would offer the possibility to call a JAX emulator of the Cosmic Linear Anisotropy Solving System \citep{2011JCAP...07..034B}.

To make the bridge with observations, two pytrees are introduced, namely \texttt{NumberCounts} and \texttt{WeakLensing} tracers for galaxy clustering and weak lensing  respectively,  with for each probe a bunch of redshift bins. A long this line, the angular power spectra $C_\ell^{pq}(z,z^\prime)$ for the probes $(p,q)$ and for redshift bin window selections $(N(z), N(z^\prime))$ are computed in the Limber approximation. 
Remarkably, all the integral involved in the computation can be easily differentiated with respect to the cosmological parameters thanks to \textit{autodif} and the implementation of the above mentioned integration quadrature.  An example is given in the context of DES Y1 3x2pts analysis (\ref{app-DESY1}).

@@

%\subsection{Background cosmology}
% Usual stuff

%\subsection{Growth of perturbations}
% The interesting point to mention here is the differentiable growth ODE

%\subsection{Matter power spectrum}
% Here we describe the EH power spectrum, mention possible extensions with emulators

%\subsection{Tracers}
% Here we describe the kernels, like in CCL

%\subsection{Angular power-spectra}
% Limber integration, this can be an opportunity to talk about differentiation through integrals as well.

%
%\subsection{Validation against the Core Cosmology Library (CCL)}
To illustrate the different features available with the present version of the library, we have written a  notebook to make  comparison with the well established  Core Cosmology Library \citep{2019ApJS..242....2C}. As an example, the Figure \ref{fig:halofit_comparison} shows an example of the non linear matter power spectrum computation using the \texttt{halofit} fitting function \citep{2012ApJ...761..152T, 2003MNRAS.341.1311S}. 
%
\begin{figure}[ht]
    \centering
    \includegraphics[width=\columnwidth]{figures/halofit_pk.png}
    \caption{Comparison of matter power spectrum between CCL and jax-cosmo. \textbf{Example of validation plot} \nblink{CCL_comparison}}
    \label{fig:halofit_comparison}
\end{figure}

@@

%
%
\subsection{DES-Y1 Example}
%
%\Acknowledgements
\section*{Acknowledgements}
Some of the numerical experiments have been conducted at the IN2P3 Computing Center (CC-IN2P3 - Lyon/Villeurbanne - France) funded by the Centre National de la Recherche Scientifique.

@@

Then, using the \numpyro\ PPL, we have set up a \textit{forward model} using the prior description  Table \ref{tab-DESY1} as for instance the $\Omega_c$ parameter
\begin{minted}[fontsize=\footnotesize]{python}
Omega_c = numpyro.sample('Omega_c', Uniform(0.1, 0.9))
\end{minted}
Then, we generate mock angular power spectra $C_\ell$ from the auto \& cross correlations of the Number Counts and Weak Lensing probes computed with the \texttt{gaussian\_cl\_covariance\_and\_mean} function in the \texttt{jc.angular\_cl} module. One reads
\begin{minted}[fontsize=\footnotesize]{python}
# Define the lensing and number counts probe
    probes = [jc.probes.WeakLensing(nzs_s_sys, 
                                    ia_bias=b_ia,
                       multiplicative_bias=m),
             jc.probes.NumberCounts(nzs_l, b)]
cl, C = gaussian_cl_covariance_and_mean(cosmo, ell, probes, 
                               f_sky=0.25, sparse=True)
\end{minted} 
with \texttt{cosmo} an instance of the \texttt{jc.Cosmology} setting the cosmological parameters generated with the priors, and \texttt{ell} a series of 50 angular modes. The generation of the mock data yields from the possibility to execute a conditioned model with a fiducial cosmology, leaving \numpyro\ dealing with the random number generation:
\begin{minted}[fontsize=\footnotesize]{python}
fiducial_model = condition(model,
    {'Omega_c':0.2545, 'sigma8':0.801, 
    'h':0.682, 'Omega_b':0.0485, 'w0':-1.,'n_s':0.971,
     'A':0.5,'eta':0.,
     'm1':0.0,'m2':0.0,'m3':0.0,'m4':0.0,
     'dz1':0.0,'dz2':0.0,'dz3':0.0,'dz4':0.0,
     'b1':1.2,'b2':1.4,'b3':1.6,'b4':1.8,'b5':2.0
      })
